Alien species

### 2023-03-22 DRAFT_alien_species.Rmd
types etc

```{r, echo=F}
Ecosystem <- "Våtmark, Naturlig åpne områder under skoggrensa, Semi-naturlig mark" 
Egenskap  <- "Primærproduksjon (evt. Abiotiske fohold)"
ECT       <- "Physical state characteristics" 
Contact   <- "anders.kolstad@nina.no"
```

@@

## Introduction
This indicator represent the the ecological on-site effect from  alien species. Whilst the introduction of alien species is best considered a pressure variable, the physical abundance or relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a good condition indicator.

@@

This time we are using the [nature type mapping dataset](#naturtype) which records the _effect_ of alien species,, rather than the % cover. The recording scale is called R7 (Fig. \@ref(fig:R7)) and is an ordinal scale which has a non-linear relationship with %cover recorded in ANO (Fig. \@ref(fig:fremmedart-prosentVSinnslag)). In order to combine these two data sets, we must convert one of these scales to match the other.

@@

```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with %cover (y-axis), which is how alien species is recorded in ANO. I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here exualrs to an R7 value of 1. Cropped from Evju et al (2022)."}
knitr::include_graphics("images/fremmedart_prosentVSinnslag.PNG")
```

Since the nature type data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space. We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure, and we say that the field data is representative inside the HIA and region. In [a related indicator](#slitasje-og-kjørespor) we used the accounting areas as regions (5 accounting areas in Norway). For this indicator, we expect more gradients also inside accounting areas, and therefore we define areas as municipalities. This means that there will be more missing data in our final indicator maps, since not all municipalities will have sufficient data points to estimate indicator values.

We then calculate an area weighted mean (and error) indicator value for each HIA and municipality combination, as long as there is more than _n_ data points for a given combination of HIA and municipality.

@@

The indicator uses a data set from a standardised field survey of nature
types. You can read more about the preliminary analyses
[here](#naturtype). See also the [official
site](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/naturkartlegging/naturtyper/)
of the Environment Agency. I also import a data set called
[ANO](#ANO-import2), which you can read about
[here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/)

@@

The variables are recorded on a seven-step ordinal scale (Fig.
\@ref(fig:R7)) or a %cover of alien species.

@@

The indicator is tagged to the *Økologisk egenskap* called
**Funksjonelt viktige arter og strukturer** (Functionally important species and structures).

@@

We also have a separate [summary file](##exp-natureType-summary) where
the nature types are manually mapped the NiN variables and to the correct NiN-main types. 
We can use this to find the nature types associated with the NiN main types
that we are interested in.

@@

Importing and sub-setting the main data file, fix duplicate _hovedøkosystem_, calculate area, split one column in two, make numeric, and select target variables:
```{r, message=F}
naturetypes <- sf::st_read(dsn = path) %>%
  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
  mutate(hovedøkosystem = case_match(hovedøkosystem,
                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
                                  .default = hovedøkosystem),
         area = st_area(.)) %>%
  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  filter(NiN_variable_code %in% myVars)
```

@@

```{r alien-naturetype, fig.cap="Alien species variable scores."}
ggplot(naturetypes, aes(x = NiN_variable_value))+
  geom_histogram(fill="grey",
           colour = "black",
           binwidth = 1)+
  theme_bw(base_size = 12)+
  labs(x = "Alien species variable score") +
  facet_wrap(.~hovedøkosystem, scales="free_y")
```

@@

_NOTE: I'm not sure, but I wonder if the 2021 GRUK data was a pilot project and that most GRUK data, and all futute GRUK data, will not be included in this data set._

@@

Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
placed locations each with 18 sample points. In each sample point a
circle of 250 m2 is visualised, and the main ecosystem is recorded.

@@

It appears the proportion of each circle that is made up of the dominant
ecosystem was only recorded after year 2019. In fact, the main ecosystem
was not recorded at all in 2019:
```{r}
# Hard coding name change
#unique(ano$hovedtype_250m2)[3]
#ano$hovedtype_250m2[ano$hovedtype_250m2==unique(ano$hovedtype_250m2)[3]] <- "Aapen jordvassmyr"
```

@@

The zero in there is an obvious mistake.

@@

```{r ano-alien-dist, fig.cap"Distribution of %cover of alien plants variable in the ANO data set"}
ggplot(ano, aes(x = fa_total_dekning))+
  geom_histogram(fill="grey",
           colour = "black",
           binwidth = 1)+
  theme_bw(base_size = 12)+
  labs(x = "% cover alien vascular plants") +
  facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
```

@@

Now I will convert these %cover values into 7FA classes. This is a critical step, since there is not one-to-one mapping here, 
but rather there is considerable overlap, as can be seen in Fig. \@ref(fig:fremmedart-prosentVSinnslag).

@@

```{r ano-alien-dist-R7, fig.cap"Distribution of R7 values in the ANO data set"}
ggplot(ano, aes(x = var_7FA))+
  geom_histogram(fill="grey",
           colour = "black",
           binwidth = 1)+
  theme_bw(base_size = 12)+
  labs(x = "% cover alien vascular plants\n converted to the R7 scale") +
  facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
```

@@

Because later we might want to save this as a shape file, we cannot have
a mixed class type. I will therefore convert the points to polygons. I
use the area column to calculate a radii that gives that area.

@@

Let's see the proportion of data points (not area) origination from each
data set

```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the alien species indicator.", fig.width=4}
temp <- as.data.frame(table(alien_data$origin))

@@

The GIS layers are used to crop out marine areas, and to define
accounting areas, respectively.

@@

However, the combined surveyed area is a very small fraction of the
total area of Norway, so that only producing indicator values for the
mapped areas leaves the indicator without much value for regional
assessments. When we do regional assessments and calculate regional
indicator values, we cannot simply do an area weighting of the polygons
in each region. This is because we don't want to assume that the
polygons are representative far outside of the mapped area. But perhaps
we can assume them to be representative inside *homogeneous ecological
areas,* or *Homoegeneous Impact Areas* (HIAs). That's where the
infrastructure index comes in. Here's the plan:

@@

2.  take a simplified infrastructure index (vector data set with HIAs)
    and extract the corresponding indicator values that intersect with a
    given HIA, and extrapolate the area weighted mean of those values to
    the entire HIA withing a given geographical region (municipality). Errors are
    calculated from bootstrapping.

3.  Take the new area weighted indicator values for the HIAs and match
    them with ecosystem occurrences of the same HIA (_possibly using
    randomly sampled points from the available ecosystem occurrences_).
    The errors are the same as for the HIAs.

>>>>

```{r, echo=F}
Ecosystem <- "Våtmark, Naturlig åpne områder under skoggrensa, Semi-naturlig mark" 
Egenskap  <- "Funksjonelt viktige arter og strukturer"
ECT       <- "B1 Compositional state characteristics" 
Contact   <- "anders.kolstad@nina.no"
```

@@

## Introduction
This indicator represent the the ecological on-site effect from  alien species. Whilst the introduction of alien species is best considered a pressure variable, the relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a valid condition indicator.

@@

This time we are using the [nature type mapping dataset](#naturtype) which records the assumed _effect_ of alien species, rather than a direct measure of abundance such as % cover. The recording scale is called R7 (Fig. \@ref(fig:R7)) and is an ordinal scale which has a non-linear relationship with %cover recorded in ANO (Fig. \@ref(fig:fremmedart-prosentVSinnslag)). In order to combine these two data sets, we must convert one of these scales to match the other.

@@

```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with %cover (y-axis), which is how alien species is recorded in ANO. I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here equals an R7 value of 1. Cropped from Evju et al (2022)."}
knitr::include_graphics("images/fremmedart_prosentVSinnslag.PNG")
```

Since the nature type data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space. We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure intersected with municipality borders, and we say that if we have more than _n_ data points than this field data is representative inside the entire HIA. In [a related indicator](#slitasje-og-kjørespor) we used the accounting areas to define the HIA (5 accounting areas in Norway), in addition to the infrastructure classes. For this indicator, we expect more gradients also inside accounting areas, and therefore we define HIAs based on municipalities. This means that there will be more missing data in our final indicator maps, since not all municipalities will have sufficient data points to estimate indicator values.

We then calculate an area weighted mean (and error) indicator value for each HIA, as long as there is more than _n_ data points for a given combination of HIA and municipality.

@@

The indicator uses a data set from a standardised field survey of nature
types. You can read more about this data set from my preliminary analyses
[here](#naturtype). See also the [official
site](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/naturkartlegging/naturtyper/)
of the Environment Agency. I also import a data set called
[ANO](#ANO-import2), which you can read about
[here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/).

@@

The variables are recorded on a unitless seven-step ordinal scale (Fig.
\@ref(fig:R7)) or a as a % cover of alien species.

@@

The indicator is tagged to the *Økologisk egenskap* called
**Funksjonelt viktige arter og strukturer** (Functionally important species and structures). This emphasises the negaivte effect that alien species has on for example supressing or excluding native species. In some cases, alien species can also affect nutrient cycling or recruitment.

@@

We also have a separate [summary file](##exp-natureType-summary) where
the nature types are manually mapped to the NiN variables and to the correct NiN-main types. 
We can use this to find the nature types of interest.

@@

Importing and sub-setting the main data file, fix duplicate _hovedøkosystem_, calculate area, split one column in two, make numeric, and select target variables:
```{r, message=F, warning=FALSE}
naturetypes <- sf::st_read(dsn = path) %>%
  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
  mutate(hovedøkosystem = case_match(hovedøkosystem,
                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
                                  .default = hovedøkosystem),
         area = st_area(.)) %>%
  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  filter(NiN_variable_code %in% myVars)
```

@@

```{r alien-naturetype, fig.cap="Alien species variable scores in the nature type data set."}
ggplot(naturetypes, aes(x = NiN_variable_value))+
  geom_histogram(fill="grey",
           colour = "black",
           binwidth = 1)+
  theme_bw(base_size = 12)+
  labs(x = "Alien species variable score") +
  facet_wrap(.~hovedøkosystem, scales="free_y")
```

@@

_NOTE: I'm not sure, but I wonder if the 2021 GRUK data was a pilot project and that most GRUK data, and all future GRUK data, will not be included in this data set._

@@

Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
placed locations in Norway, each with 18 sample points. In each sample point a
circle of 250 m2 is visualised, and the main ecosystem is recorded.

@@

It appears the proportion of each circle that is made up of the dominant
ecosystem was only recorded after year 2019. In fact, the main ecosystem
was not recorded at all in 2019:

@@

The zero in there is an obvious mistake. Removing it:

@@

```{r ano-alien-dist, fig.cap="Distribution of %cover of alien plants variable in the ANO data set"}
ggplot(ano, aes(x = fa_total_dekning))+
  geom_histogram(fill="grey",
           colour = "black",
           binwidth = 1)+
  theme_bw(base_size = 12)+
  labs(x = "% cover alien vascular plants") +
  facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
```

@@

Now I will convert these %cover values into 7FA classes. This is a critical step, since there is no one-to-one mapping here, 
but rather there is considerable overlap, as can be seen in Fig. \@ref(fig:fremmedart-prosentVSinnslag).

@@

```{r ano-alien-dist-R7, fig.cap="Distribution of R7 values in the ANO data set"}
ggplot(ano, aes(x = var_7FA))+
  geom_histogram(fill="grey",
           colour = "black",
           binwidth = 1)+
  theme_bw(base_size = 12)+
  labs(x = "% cover alien vascular plants\n converted to the R7 scale") +
  facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
```

@@

Because later we will rasterize these data, we will  convert the points to polygons. I
use the area column to calculate a radii that gives that area.

@@

Let's see the proportion of data points (not area) originating from each
data set

```{r, fig.cap="Barplot showing the contribution (number of localities) of different data sets to the alien species indicator.", fig.width=4}
temp <- as.data.frame(table(alien_data$origin))

@@

These layers are used to crop out marine areas, and to define
accounting areas, respectively.

@@

However, the combined surveyed area is a very small fraction of the
total area of Norway, so that only producing indicator values for the
mapped areas leaves the indicator without much value for regional
assessments. But we cannot simply do an area weighting of the polygons
in each region. This is because we expect considerable sampling bias
and we can't assume that the
polygons are representative far outside of the mapped area. But perhaps
we can assume them to be representative inside *homogeneous ecological
areas,* or *Homoegeneous Impact Areas* (HIAs). That's where the
infrastructure index comes in. Here's the plan:

@@

2.  take a simplified infrastructure index (vector data) and intersect this with municipality boarders to produce a map of homogeneous impact areas (HIAs).

1.  extract the corresponding indicator values that intersect with a
    given HIA, and extrapolate the area weighted mean of those values to
    the entire HIA. Errors are
    calculated from bootstrapping.

3.  Take the new area weighted indicator values for the HIAs, rasterize it, and mask these
    with ecosystem occurrences. The errors are the same as for the HIAs.
